% !TEX root = ./notes.tex
\chapter[Equation of State]{The Equation of State}\label{ch.equation-of-state}

In statistical equilibrium, we can describe a system of particles by a distribution function $f(\vp,\vx)\,\dif^{3}p\,\dif^{3}x$, such that the number of particles is
\begin{equation}\label{e.dist}
N = \int \dif^{3}p\,\dif^{3}x\nsp f(\vp,\vx),
\end{equation}
where the integration is over the phase spaces of momentum and position coordinates $(\vp,\vx)$. In an ideal gas, the particles do not interact. In such a case, the distribution function $f = f(\vp)$ does not depend on position.  The integration over $\dif^{3}x$ just gives a factor of the volume, so the number density is $n = \int\dif^{3}p f(vp)$.  From equation~(\ref{e.dist}), we can get our other thermodynamic quantities: for example.
\begin{eqnarray}
\frac{E}{V} \equiv u = \int  \varepsilon(\vp) f(\vp)\,d^{3}p &\qquad& \textrm{energy per unit volume;}\label{e.u}\\
P = \int  (\vp\vdot \bvec{e}_{z} )(\bvec{v}\vdot\bvec{e}_{z}) f(\vp)\,d^{3}p &\qquad& \textrm{pressure}.\label{e.P}
\end{eqnarray}
Here $\varepsilon$ is the particle energy and $v$ the velocity.

\section{An ideal Fermi gas}

For \emph{fermions}, particles with half-integer spin, it can be shown that
\begin{equation}\label{e.fermion}
f(p) = \frac{g}{(2\pi\hbar)^{3}} \left[\exp\left(\frac{\varepsilon-\mu}{\kB T}\right)+1\right]^{-1}.
\end{equation}
In this equation $\varepsilon(p)$ is the energy of a particle, $\mu$ is the \emph{chemical potential}, $T$ is the temperature, and $g$ denotes the number of particles that can occupy the same energy level (for spin-1/2 particles, $g=2$).  The connection to thermodynamics is via the relations
\[
\frac{1}{T} = \left(\frac{\partial S}{\partial E}\right)_{N,V},\qquad -\frac{\mu}{T} = \left(\frac{\partial S}{\partial N}\right)_{E,V},
\]
and 
\[
TS = PV - \mu N + E;
\]
these are derived in standard texts. Let's explore what happens in various limits.  

\subsection{Non-degenerate, non-relativistic limit}
First, let's take $K\equiv\exp(\mu/\kB T)\ll 1$. (In the literature, $K$ is called the \emph{fugacity}.) Then in equation~(\ref{e.fermion}), we see that the exponential term dominates.  If our system is isotropic, then $d^{3}p = 4\pi p^{2}dp$, and we'll use this substitution from now on.  We then have for the number density
\begin{equation}
n(\mu,T) = \frac{gK}{2\pi^{2}\hbar^{3}}\int_{0}^{\infty}\exp\left(-\frac{\varepsilon}{\kB T}\right) p^{2}dp.
\end{equation}
To do the integral, notice that since $2m\varepsilon = p^{2}$, we have $p^{2}\,dp = m(2m\varepsilon)^{1/2}\,d\varepsilon$; making the substitution $x = \varepsilon/(\kB T)$, we get
\begin{equation}\label{e.n-int}
n(\mu,T) = \frac{gK}{2\pi^{2}\hbar^{3}}\sqrt{2}(m\kB T)^{3/2}\int_{0}^{\infty} x^{1/2}e^{-x}\,dx.
\end{equation}
You have all struggled with this integral in your past, but to avoid unpleasant flashbacks, I will just tell you that it is $\sqrt{\pi}/2$.  So, we have our first result (but we still don't know what it means),
\begin{equation}\label{e.n-nondeg}
n(\mu,T) = K\left[g\left(\frac{m\kB T}{2\pi\hbar^{2}}\right)^{3/2}\right].
\end{equation}

Let's forge on a little further, though, and try to get the energy per unit voume $u$.  Once we have $u$, we know we can get the pressure from the relation for a non-relativistic gas, $P = 2/3\;u$.  Using equations~(\ref{e.u}) and (\ref{e.fermion}),
\begin{equation}
u(\mu,T) = \frac{gK}{2\pi^{2}\hbar^{3}}\int_{0}^{\infty}\exp\left(-\frac{\varepsilon}{\kB T}\right) \varepsilon p^{2}dp.
\end{equation}
Let's repeat our trick of changing variables from $p$ to $x = \varepsilon(p)/\kB T$; we then have
\begin{equation}
u(\mu,T) = \frac{gK}{2\pi^{2}\hbar^{3}}\sqrt{2}m(m\kB T)^{3/2}\int_{0}^{\infty} x^{3/2}e^{-x}\,dx.
\end{equation}
Did you notice that if we integrate by parts,
\[
\int_{0}^{\infty} x^{3/2}e^{-x}\,dx =  \frac{3}{2}\int_{0}^{\infty}x^{1/2}e^{-x}\,dx = \frac{3}{4}\sqrt{\pi},
\]
we get the integral we already solved in equation~(\ref{e.n-int})?  Putting everything together and using the expression for $n$ (eq.~[\ref{e.n-nondeg}]), we have
\begin{equation}
u(\mu,T) = \frac{3}{2}\left\{K\left[g\left(\frac{m\kB T}{2\pi\hbar^{2}}\right)^{3/2}\right] \right\}\kB T = \frac{3}{2}n\kB T.
\end{equation}
which gives us the pressure,
\begin{equation}\label{e.PVnkT}
P = \frac{2}{3} u = n \kB T.
\end{equation}
Whoo-hoo!  We've rediscovered the ideal gas.

Now we have to understand this chemical potential $\mu$.  We can solve equation~(\ref{e.n-nondeg}) for $\mu$,
\begin{equation}\label{e.fugacity-nondeg}
\exp\left(\frac{\mu}{\kB T}\right) = K = n\left[g\left(\frac{m\kB T}{2\pi\hbar^{2}}\right)^{3/2}\right]^{-1}.
\end{equation}
Now $K$ is dimensionless, a number, so the thing in $[\;]$ must have dimensions of number density.  Let's call it $n_{Q}$.  Our chemical potential is then $\mu = \kB T\ln(n/n_{Q})$.  To understand the significance of $n_{Q}$, let's calculate the uncertainty in position of a particle having energy $\kB T$; from Heisenberg, we have
\[
\Delta x \approx \frac{\hbar}{\Delta p} \sim \frac{\hbar}{\sqrt{m\kB T}}
\]
where I am dropping numerical factors and I've made the substitution $\Delta p\sim p \approx \sqrt{m\kB T}$.  Now what happens if I pack the particles so that on average there are $g$ particles per box of volume $(\Delta x)^{3}$?  In that case the density would be $n = g(\sqrt{m\kB T}/\hbar)^{3} \approx n_{Q}$.  So, what appears in the chemical potential is the ratio of the density to that density at which the particles are packed so closely that the uncertainty in their positions is the same size as the typical inter-particle spacing.
In the ideal-gas limit $K \ll 1$, which makes sense: $n\ll n_{Q}$, so the particles are very far apart compared to their thermal de Broglie wavelengths, and quantum effects ought to be unimportant.

\subsection{Degenerate, non-relativistic limit}\label{s.deg-limit-fermions}
When $n\gtrsim n_{Q}$, we can no longer use the approximation $K \ll 1$, so let's go to the opposite limit, for which $\mu \gg \kB T$.  In this case, notice from equation~(\ref{e.fermion}) that
\begin{equation}
f(p) \approx \frac{g}{(2\pi\hbar)^{3}}\left\{\begin{array}{lr} 1 & \varepsilon < \mu\\ 0 & \varepsilon> \mu\end{array}\right. .
\end{equation}
We can think of this as inserting $g$ particles in each energy level, starting with the lowest energy level and continuing until all of the particles are used. The last particle is inserted with energy $\varepsilon \approx \mu$.  The only levels that will be partially filled will be those lying in a thin band $\varepsilon \approx\mu \pm \kB T$.  If that is the case, we can make the following approximation.  Let's take the limit $T \to 0$, and define the \emph{Fermi energy} by $\eF = \mu|_{T\to 0}$ and the \emph{Fermi momentum} by $\pF = \sqrt{2m\eF}$.  We can then write equation~(\ref{e.dist}) as
\begin{equation}
n(\mu) = \frac{1}{\pi^{2}\hbar^{3}}\int_{0}^{\pF} p^{2} dp,
\end{equation}
since the integrand is zero for $p > \pF$.  The main application is for electrons, which are spin one-half, so we substitute $g = 2$. Now this is an easy integral,
\begin{equation}\label{e.n-deg}
n(\mu) = \frac{\pF^{3}}{3\pi^{2}\hbar^{3}} = \frac{(2m\eF)^{3/2}}{3\pi^{2}\hbar^{3}},
\end{equation}
or $\mu \approx \eF = (3\pi^{2}n)^{2/3} \hbar^{2}/(2m)$.  Let's get the energy per unit volume and the pressure,
\begin{equation}
u(\mu) = \frac{1}{\pi^{2}\hbar^{3}}\int_{0}^{\pF} \frac{p^{2}}{2m}\,p^{2}\,dp = \frac{\pF^{5}}{5\pi^{2}\hbar^{3}}.
\end{equation}
Comparing this with equation~(\ref{e.n-deg}), we have
\begin{eqnarray}
u &=& \frac{3}{5}n\eF,\label{e.u-deg}\\
P &=& \frac{2}{5}n\eF\label{e.P-deg}.
\end{eqnarray}
To lowest order, neither $u$ nor $P$ depend on $T$.  Substituting for \eF\ in equation~(\ref{e.P-deg}) gives us the equation of state,
\begin{equation}\label{e.eos-deg}
P = \frac{2}{5}\left(3\pi^{2}\right)^{2/3}\frac{\hbar^{2}}{2m}n^{5/3}.
\end{equation}
Notice that in equation~(\ref{e.fugacity-nondeg}), $n_{Q}\propto m^{3/2}$.  This means that at any given temperature, $n_{Q}$ for electrons is $1836^{3/2} = 80,000$ times smaller than it is for protons, not to mention helium or heavier nuclei. As a result, the electrons will become degenerate ($n\gtrsim n_{Q}$) at a much lower mass density than the ions.    A common circumstance, then, is to have a mixture of degenerate electrons and ideal ions (we will deal with non-ideal corrections due to electric forces later).  

Now, let's estimate the boundary between the non-degenerate and degenerate regimes. At a given temperature, we know in the low-density limit that the electrons obey the ideal gas law (eq.~[\ref{e.PVnkT}]) and in the high-density limit the electrons are degenerate (eq.~[\ref{e.eos-deg}]). So, let's extrapolate our two limiting expressions for the pressure and see where they meet,
\begin{equation}
n_{e}\kB T = P_{e,\mathrm{ideal}} \sim P_{e,\mathrm{deg.}} 
  = \frac{2}{5}n_{e}\eF,
\end{equation}
or $\eF \approx \kB T$.  No surprise here.  Notice that the ratio
\begin{equation}
\frac{\eF}{\kB T} \sim \frac{(3\pi^{2})^{2/3}\hbar^{2}}{2m_{e}\kB T}n_{e}^{2/3} \sim \left(\frac{n_{e}}{n_{Q}}\right)^{2/3},
\end{equation}
so marking the onset of degeneracy with $\eF\sim \kB T$ also makes sense from that aspect as well. Our boundary in the density-temperature plane between the non-degenerate and degenerate regimes is then determined by setting $\kB T = \eF$,
\begin{equation}\label{e.TF}
T = \frac{(3\pi^{2})^{2/3}\hbar^{2}}{2m_{e}\kB }\left(\frac{Y_{e}\rho}{\mb}\right)^{2/3} = 3.0\ee{5}\nsp\K \left(Y_{e}\rho\right)^{2/3}.
\end{equation}
If the temperature falls below this value, the electrons will be degenerate. Here $Y_{e}$ is the electron molar fraction, or electron abundance and \mb\ is the atomic mass unit; consult \S\ref{s.composition} for details.

\section{Fermi-Dirac integrals}
This condition for the onset of degeneracy, eq.~(\ref{e.TF}), is only a rule-of-thumb; in any serious calculation we would want to calculate the electron thermal properties from the exact integrals
\begin{eqnarray}\label{e.FD12}
n(\mu,T) &=& \frac{\sqrt{2}(m\kB T)^{3/2}}{\pi^{2}\hbar^{3}}\int_{0}^{\infty} \frac{x^{1/2}\,dx}{\exp(x-\psi)+1} \\
P(\mu,T) &=& \frac{(2m\kB T)^{3/2}(\kB T)}{3\pi^{2}\hbar^{3}}\int_{0}^{\infty} \frac{x^{3/2}\,dx}{\exp(x-\psi)+1},
\end{eqnarray}
where $\psi = \mu/(\kB T)$.  These integrals cannot be done analytically, but they occur so frequently that there are many published tables and numerical approximation schemes \citep[see][]{timmes.swesty:accuracy}.  Specifically, the \emph{non-relativistic Fermi-Dirac integral of order $\nu$} is defined as
\begin{equation}\label{e.FDintegral}
F_{\nu}(\psi) = \int_{0}^{\infty}\frac{ x^{\nu}\,dx}{\exp(x-\psi)+1}.
\end{equation}
One can (numerically) invert equation~(\ref{e.FD12}) to solve for the chemical potential $\psi \kB T$.

The general case for a relativistic Fermi gas is left as an exercise.

\section{Relativistic photon gas}\label{s.relativistic-photon-gas}
Photons are \emph{bosons}---they have spin 1. For bosons, the distribution function is similar to that in equation~(\ref{e.fermion}), but with the $+1$ replaced by $-1$ in the denominator. In addition, photon number is not conserved: one can freely create and destroy photons.  This implies that their chemical potential is zero.  Also, $g=2$ for photons: there are two independent polarization modes. Putting all of these together, we can write energy per unit volume as
\begin{equation}\label{e.boson-dist}
u = \frac{1}{\pi^{2}\hbar^{3}}\int_{0}^{\infty}\varepsilon p^{2}\left[\exp\left(\frac{\varepsilon}{\kB T}\right)-1\right]^{-1}\,dp.
\end{equation}
Now, use the fact that $p = \varepsilon/c$ and change variables to $x = \varepsilon/(\kB T)$ to get
\[
 u = \frac{\kB ^{4}T^{4}}{\pi^{2}c^{3}\hbar^{3}}\int_{0}^{\infty} \frac{x^{3}\,dx}{e^{x}-1}.
\]
This integral is a classic and is equal to $\pi^{4}/15$.  Hence the energy per unit volume and the pressure are
\begin{eqnarray}\label{e.radiation-eos}
u &=& \left(\frac{\kB ^{4}\pi^{2}}{15 c^{3}\hbar^{3}} \right) T^{4} = aT^{4}\\
P &=& \frac{1}{3} aT^{4}.
\end{eqnarray}
In CGS units, $a = 7.566\ee{-15}\nsp\ergs\usp\cm^{-3}\usp\K^{-4}$.

\section{Connection to thermodynamics}
Once we have the distribution functions, we can get all of the other thermodynamic properties from the thermodynamic relations: in what follows let $N = nV$ be the total number of particles, with $V$ being the volume of the system.  The total energy is then $E = uV$, and the entropy is $S$, and we have
\begin{eqnarray}
F = E - TS, &\qquad& \textrm{Helmholtz free energy,}\\
H = E + PV, &\qquad& \textrm{Enthalpy,}\\
\mu N = G = A + PV, &\qquad& \textrm{Gibbs free energy}.
\end{eqnarray}
For example, we have in the non-degenerate limit that
\begin{equation}\label{e.chem-pot-ideal-gas}
\mu = \kB T\ln K = \kB T\ln\left[ \frac{n}{g}\left(\frac{2\pi\hbar^{2}}{m\kB T}\right)^{3/2}\right],
\end{equation}
and so we could write the entropy per unit mass as
\begin{eqnarray}
s \equiv \frac{S}{Nm} &=& \frac{1}{Nm}\frac{E + PV - \mu N}{T}\nonumber\\
 &=& \frac{\kB }{m}\left\{ \frac{5}{2} + \ln\left[\frac{g}{n}\left(\frac{m\kB T}{2\pi\hbar^{2}}\right)^{3/2}\right]\right\}.
\label{e.sacker-tetrode}
 \end{eqnarray}
In this equation I have used the ideal non-degenerate values $E = (3/2) N\kB T$, $PV = N\kB T$ and have denoted the mass per particle as $m$ and the degeneracy of the spin-states as $g$.

\section{Chemical Equilibrium: The Saha Equation}

Consider a reaction, $A + B + \ldots \to C + D + \ldots$. When this reaction comes into equilibrium, we are at a maximum in entropy, and the condition for equilibrium is that the energy cost, at constant entropy to run the reaction in the forward direction is the same as to run the reaction in reverse. This can be expressed in terms of chemical potentials as
\begin{equation}\label{e.mass-action}
\mu_{A} + \mu_{B} + \ldots \to \mu_{C} + \mu_{D} + \ldots
\end{equation}
Note in this formalism that a reaction $2A \to B$ would be expressed as $2\mu_{A} = \mu_{B}$.

As a worked example, we consider the ionization equilibrium of hydrogen,
\[ \mathrm{H^{+}} + e \to \mathrm{H}. \]
To use equation~(\ref{e.mass-action}), we need to have both sides on the same energy scale. The reaction in the exothermic direction; that is, heat is evolved if the reaction proceeds as written.  This means that the right-hand side is more bound, and its minimum energy is less than that of the right-hand side. To get both sides on the same energy scale, we must subtract the binding energy, $Q=13.6\nsp\eV$, from the right-hand side:
\begin{equation}\label{e.ionization-chem-potential}
\mu_{+} + \mu_{-}= \mu_{0} -  Q.
\end{equation}
Another way to see why $Q$ appears is to add the rest mass for each species to its chemical potential; collecting all terms on the right, we would then have a term $(m_{0}-m_{+}-m_{-})c^{2} = -Q$. Some ionization potentials, along with the half-ionization temperature for a given electron number density $n_{e}$, are given in Table~\ref{t.ionization-potentials}.

\begin{table}[htpb]
\begin{center}
\caption{Selected ionization potentials and half-ionization temperatures}\label{t.ionization-potentials}
\begin{tabular}{l|rrrr}
\hline
element & H$^{-}$ & Na & H & He\\
\hline\hline
$Q/(\mathrm{eV})$ & 0.75 & 5.14 & 13.6 & 24.6\\
$T_{1/2}(n_{e} = 10^{13}\nsp\cm^{-3})/\mathrm{K}$ & 600 & 3300 & 8\,000 & 13\,900\\
$T_{1/2}(n_{e} = 10^{16}\nsp\cm^{-3})/\mathrm{K}$ & 900 & 5\,000 & 11\,800 & 20\,200\\
\hline
\end{tabular}
\end{center}
\end{table}

For a non-degenerate plasma, we can insert eq.~(\ref{e.chem-pot-ideal-gas}) into eq.~(\ref{e.ionization-chem-potential}), divide through by $\kB T$, and take the exponential to obtain
\begin{equation}\label{e.saha-eqn}
\frac{n_{+}n_{-}}{n_{0}} = \frac{g_{+}g_{-}}{g_{0}}\left(\frac{m_{-}\kB T}{2\pi\hbar^{2}}\right)^{3/2}\exp\left(-\frac{Q}{\kB T}\right).
\end{equation}
The number density of all hydrogen in the gas is $n_{0}+n_{+} = n_{\mathrm{H}}$.  Denote the ionized fraction by $x = n_{+}/n_{\mathrm{H}} = n_{i}/n_{\mathrm{H}}$, so that the left-hand side of equation~(\ref{e.saha-eqn}) is $n_{\mathrm{H}} x^{2}/(1-x)$. In the hydrogen atom ground state, the electron spin and proton spin are either aligned or anti-aligned. These states are very nearly degenerate, so that $g_{0} = 2$.  Both the proton and electron have spin $1/2$; there are really only two available states, however, because of the freedom in choosing our coordinate system.  As a result, $g_{+}g_{-} = 2$ as well.

Inserting these factors into equation~(\ref{e.saha-eqn}), and using $\kB = 8.6173\ee{-5}\nsp\eV/\K$, we obtain
\begin{equation}\label{e.saha-reduced}
\frac{x^{2}}{1-x} = \frac{2.41\ee{21}\nsp\cm^{-3}}{n_{\mathrm{H}}} \left(\frac{T}{10^{4}\nsp\K}\right)^{3/2}\exp\left(-\frac{15.78\ee{4}\nsp\K}{T}\right).
\end{equation}
This equation defines a set of points in the $\rho-T$ plane for which $x = 1/2$.  We may take this set of points to mark the boundary between neutral and ionized hydrogen. At fixed density, the transition from neutral to fully ionized is very rapid.

\section{Coulomb interactions}
Under the conditions in a stellar interior, most of the atoms are ionized, and the stellar matter consists of positively charged nuclei and ions and negatively charged electrons. A \emph{plamsa} is defined as a gas of charged particles in which the kinetic energy of a typical particle is much greater than the potential energy due to its nearest neighbors. To make this quantitative, consider a gas with only one species present, with charge $q$.  Let the mean spacing between particles be $a$; clearly the number density of such particles is $n = (4\pi a^{3}/3)^{-1}$.  We may then take the quantity
\begin{equation}\label{e.Gamma-definition}
\Gamma \equiv \frac{q^{2}}{a\kB T}
\end{equation}
as indicating the relative importance of potential to kinetic energy.  In a classical plasma, $\Gamma \ll 1$.  Note, however, that systems with $\Gamma > 1$ are often (confusingly) called  \emph{strongly coupled plasmas}. The meaning is usually clear from context.

\subsection{Debye shielding}\label{s.plasma-shielding}

Imagine a typical charged particle in a plasma.  Very close to the particle, we expect the electrostatic potential to be that of an isolated charge $\Phi = q/r$. Far from the particle, there will be many other particles surrounding it, and the potential is screened. For example, a positive ion will tend to attract electrons to be somewhat, on average, closer to it than other ions: we say that the ion \emph{polarizes} the plasma.  As a result of this polarization, the potential of any particular ion should go to zero much faster than $1/r$ due to the ``screening'' from the enhanced density of opposite charges around it.

Let's consider a plasma having many ion species, each with charge $Z_{i}$, and elections. About any selected ion $j$,  particles will arrange themselves according to Boltzmann's law,
\begin{equation}\label{e.ion-boltzmann}
n_{i}(r) = n_{i0}\exp\left[-\frac{Z_{i}e\Phi(r)}{\kB T}\right].
\end{equation}
Here $n_{i0}$ is the density of particle $i$ far from the charge $j$, and $r$ is the distance between particles $i$ and $j$.  (A similar equation holds for the electrons, with $Z$ replaced by $-1$.) To solve for the potential, we can use Poisson's equation,
\begin{equation}\label{e.Poisson}
\nabla^{2}\Phi = -4\pi \sum_{i} Z_{i}e n_{i}(r) +4\pi e n_{e}(r).
\end{equation}
Our assumption is that the term in the exponential of equation~(\ref{e.ion-boltzmann}) is small, so we may expand it to first order in $\Phi$ and substitute that expansion into equation~(\ref{e.Poisson}) to obtain in spherical geometry
\[
 \frac{1}{r}\frac{\partial^{2}}{\partial r^{2}}\left(r\Phi\right) = -4\pi e \left[\sum_{i} n_{i0}Z_{i}\left(1-\frac{Z_{i}e\Phi}{\kB T}\right) - n_{e0}\left(1 + \frac{e\Phi}{\kB T}\right)\right].
\]
The overall charge neutrality of the plasma implies that $n_{e0} = \sum_{i}Z_{i}n_{i0}$; using this to simplify the above equation gives
\begin{equation}\label{e.linearized-Poisson}
\frac{1}{r}\frac{\partial^{2}}{\partial r^{2}}\left(r\Phi\right) = \left[\frac{4\pi e^{2}}{\kB T}\sum_{i} n_{i0}\left(Z_{i}^{2}+Z_{i}\right)\right]\Phi \equiv \lambdaD^{-2}\Phi.
\end{equation}
The quantity in $[\, ]$ has dimensions of reciprocal length squared and we define it as $(1/\lambdaD)^{2}$ with $\lambdaD$ being called the \emph{Debye length}.

Multiplying equation~(\ref{e.linearized-Poisson}) by $r$, integrating twice, and determining the constant of integration from the condition that as $r\to 0$, $\Phi \to Z_{j}e/r$ gives the self-consistent potential
\begin{equation}\label{e.screened-potential}
\Phi = \frac{Z_{j}e}{r}\exp\left(-\frac{r}{\lambdaD}\right).
\end{equation}
The Debye length \lambdaD\ determines the size of the screening cloud around the ion.

In order for the above derivation to be valid, we require that $\lambdaD \gg a$, where $a$ is the mean ion spacing; otherwise, there won't be any charges in our cloud to screen the potential! Equivalently, we require the number of particles in a sphere of radius \lambdaD\ to be large,
\begin{equation}\label{e.plasma-parameter}
\frac{4\pi}{3}\lambdaD^{3} \sum_{i} n_{i} \gg 1.
\end{equation}
This condition must hold if we are to treat the gas as a plasma. In the exercises you are asked to show that this condition is equivalent to $\Gamma \ll 1$.

\subsection{Corrections to the ideal gas EOS}\label{s.plasma-corrections}

In a plasma the particles are not independent: shake one particle and other nearby particles will shake as well. In statistical mechanics, this requires introducing \emph{correlation functions} to derive the equation of state.  We'll adopt the more intuitive approach of Debye and H\"uckel to get the lowest-order correction to the ideal gas EOS.  First, the total electrostatic energy in a volume $V$ is
\begin{equation}\label{e.electrostatic-energy}
E_{\mathrm{Coul}} = \frac{1}{2}V\sum_{j}Z_{j}e n_{j}\Phi_{j}.
\end{equation}
Here $\Phi_{i}$ is the potential at a particle $j$ \emph{due to all the other particles in the plasma.}  Now, we computed the total potential around a particle (eq.~[\ref{e.screened-potential}]); expanding and subtracting off the self-potential of particle $j$ gives $\Phi_{j} = -Z_{j}e/\lambdaD$.  Inserting this into equation~(\ref{e.electrostatic-energy}) and expanding gives
\begin{equation}\label{e.electrostatic-energy-2}
E_{\mathrm{Coul}} \approx - V\left(\frac{\pi}{\kB T}\right)^{1/2} e^{3} \left[\sum_{i}n_{i0} \left(Z_{i}^{2} + Z_{i}\right)\right]^{3/2}.
\end{equation}
This energy is to be added to the kinetic energy of the gas.  The effect of the electrostatic interactions is to \emph{decrease} the energy in the gas, that is, to make it more bound.

We can't directly get the pressure from equation~(\ref{e.electrostatic-energy-2}) because the equation isn't in terms of $S$ and $V$ (recall that $P = -(\partial E/\partial V)_{S}$) but rather in terms of $T$ and $V$.
In order to get the pressure, we first must find the Helmholtz free energy $F$. To do this, we integrate the thermodynamical identity
\[ E = -T^{2}\left(\frac{\partial}{\partial T}\right)_{V}\left(\frac{F}{T}\right)
\] 
and then take $P = -(\partial F/\partial V)_{T,N}$ to obtain
\begin{equation}\label{e.pressure-correction}
 P_{\mathrm{Coul}} \approx -\frac{e^{3}}{3}\left(\frac{\pi}{\kB T}\right)^{1/2}\left[\frac{\left(\langle Z^{2}\rangle+\langle Z\rangle \right)\rho}{\langle A\rangle\mb}\right]^{3/2}.
\end{equation}
The effect of Coulomb interactions is to decrease the pressure below the ideal gas value.

\subsection[Strongly Coupled Plasmas]{Coulomb corrections when the electrons are degenerate}\label{s.degenerate-coulomb}

The above discussion holds only when both the electrons and ions are non-degenerate.  What happens when the electrons are degenerate?  In that case the kinetic energy is of order the Fermi energy, not the temperature.  We might think to replace $\kB T$ with \eF\ in equation~(\ref{e.Gamma-definition}). Recalling the formula for \eF\ from \S~\ref{s.deg-limit-fermions}, we have the condition for the Coulomb interactions to be weak,
\begin{equation}\label{e.condition-weak-degen}
\frac{e^{2}}{a \eF} = \left(\frac{4\pi n}{3}\right)^{1/3}\left(\frac{m_{e} e^{2}}{\hbar^{2}}\right)\frac{2}{(3\pi^{2} n)^{2/3}} < 1.
\end{equation}
Here $m_{e}$ and $n$ denote, respectively, the electron mass and number density.  Do you recognize the quantity $\hbar^{2}/(m_{e}e^{2})$?  It is the Bohr radius, \abohr.  What is \abohr\ doing in this equation?  Well, we are looking for a quantum mechanical system in which the Coulomb interaction is comparable to the non-relativistic kinetic energy.  Does that sound like any system you've seen before?

Cleaning up equation~(\ref{e.condition-weak-degen}), our condition for the electrons to be weakly interacting when degenerate is
\begin{equation}
\left(\frac{2^{5/3}}{3\pi}\right)\left( n \abohr^{3}\right)^{-1/3} < 1,
\end{equation}
or, in terms of mass density $\rho$ and electron fraction $Y_{e}$, $(Y_{e}\rho) > 0.4\nsp\grampercc$.  As the density increases, the electron gas becomes more ideal, that is, the electrostatic interaction matters less and less.  

Just to complete the discussion on electrons, what if the electrons are relativistic?  In this case, $\eF = \pF c = (3\pi^{2}n)^{1/3}\hbar c$, and
\begin{equation}\label{e.condition-weak-rel-degen}
\frac{e^{2}}{a \eF} = \left(\frac{4}{9\pi}\right)^{1/3}\left(\frac{e^{2}}{\hbar c}\right) = 3.8\ee{-3}.
\end{equation}
In this case $\eF \propto n^{1/3}$ so the density dependence cancels.  You will note the appearance of the fine structure constant $\alpha_{\mathrm{F}} = e^{2}/(\hbar c)$, as you might have expected when dealing with relativistic electrons and electrostatics.

Under astrophysical conditions, we can almost always regard degenerate electrons as being ideal.  What about the ions?  They are not usually degenerate under conditions of interest. We can get a simple expression if we go to the opposite limit, in which the electrons are very degenerate.  In that case, the electrons are an ideal gas and hence have uniform density. If the temperature is low enough, the ions will have $Z^{2}e^{2}/(a\kB T) \gg 1$; in this case we might expect the ions will arrange themselves into a lattice that maximizes the inter-ionic spacing.

To get an estimate of the electrostatic energy, let's compute the energy of a charge-neutral sphere centered on a particular ion of charge $Z_{i}$.  Because the electrons have a uniform density, $Y_{e}\rho/\mb$, we can find the radius of the sphere $a$ by requiring it to have $Z_{i}$ electrons,
\begin{equation}\label{e.radius-ion-sphere}
\frac{4\pi}{3}a^{3}\left(Y_{e}\frac{\rho}{\mb}\right) = Z_{i},
\end{equation}
or $a = [3 Z_{i} \mb/(4\pi Y_{e}\rho)]^{1/3}$.  The potential energy of this sphere has two components. The first is due to electron-electron interactions,
\begin{equation}\label{e.ee-energy}
E_{ee} = \int_{0}^{a} \frac{q(r)\, \dif q}{r} = \frac{3}{5}\frac{Z_{i}^{2}e^{2}}{a},
\end{equation}
where $q(a) = Z_{i}e(r/a)^{3}$ is the charge in a sphere of radius $r < a$. The second component of the potential energy is due to the ion-electron interaction,
\begin{equation}\label{e.ei-energy}
E_{ei} = -Z_{i}e\int_{0}^{a}\frac{\dif q}{r} = -\frac{3}{2}\frac{Z_{i}^{2}e^{2}}{a}.
\end{equation}
Combining equations~(\ref{e.ee-energy}), (\ref{e.ei-energy}), and (\ref{e.radius-ion-sphere}) gives the total electrostatic energy for a single ion-sphere,
\begin{equation}\label{e.madelung-sphere}
E = -\frac{9}{10} \frac{Z_{i}^{2}e^{2}}{a} = -\frac{9}{10} Z_{i}^{5/3} e^{2} \left(\frac{4\pi}{3}\frac{Y_{e}\rho}{\mb}\right)^{1/3}.
\end{equation}
Multiplying this by $n_{i} = Y_{i}\rho/\mb$, summing over all ion species $i$, and defining $\langle Z^{5/3}\rangle = n^{-1}\sum Z_{i}^{5/3}n_{i}$ where $n = \sum n_{i}$, gives the total Coulomb energy per volume,
\begin{equation}\label{e.madelung-total}
E_{\mathrm{Coul}} = -\frac{9}{10} n\kB T \left[\frac{\langle Z^{5/3}\rangle e^{2}}{\kB T}\left(\frac{4\pi}{3}\frac{Y_{e} \rho}{\mb}\right)^{1/3}\right].
\end{equation}
Notice that the quantity in $[\nsp]$ reduces to $\Gamma$ for a single-species plasma.  If we therefore define $\Gamma \equiv [\nsp]$ for a multi-component plasma, we have $E_{\mathrm{Coul}} \approx -0.9 n\kB T \Gamma$; the pressure is then $P_{\mathrm{Coul}} = E_{\mathrm{Coul}}/3  = -0.3 n \kB T \Gamma$. This holds in the limit $\Gamma \gg 1$.

\section{Exercises}\label{s.EOS-exercises}
\begin{enumerate}
\item For an isotropic momentum distribution, show that 
\[
P = \frac{1}{3}\int |p||v| f(\vp)\,\dif^{3}\vp.
\]
\item Now show that for a non-relativistic gas, $P = (2/3)E/V$, and that for a relativistic gas, $P = (1/3)E/V$.

\item Repeat the derivation of equation~(\ref{e.u-deg}) and (\ref{e.P-deg}) for  a relativistic Fermi gas. What is the expression for the temperature at which the gas becomes degenerate (cf.~eq.~[\ref{e.TF}]) in this case?

\item Get the first order corrections to the Maxwell-Boltzmann gas. Take the fugacity $K \ll 1$, and expand the Fermi-Dirac distribution  (eq.~[\ref{e.fermion}]) to lowest order in $K\exp[-\varepsilon/(\kB T)]$. Show that 
\begin{eqnarray*}
 n(K,T) &=& n_{0}\left(1 - 2^{-3/2}K\right) \\
 u(K,T) &=& \frac{3}{2}n_{0}\kB T \left(1-2^{-5/2}K\right),
\end{eqnarray*}
where $n_{0}$ is the density in the limit $K\to 0$.
Then derive the equation of state $P = P(n,T)$ to lowest order in $K$. For a given ($n, T$), is the pressure larger or smaller than that of the ideal Maxwell-Boltzmann limit?

\item In an external field (i.e., gravitational) the chemical potential, which is the change in energy when the number of particles is increased, must include the potential.  Consider an ideal gas in a planar atmosphere of  constant gravitational acceleration $g$.  Write the chemical potential as $\mu(z)  = \mu_{\mathrm{id}} + \Phi$, where $\mu_{\mathrm{id}}$ is the chemical potential for an ideal gas in the absence of gravity, and $\Phi$ is the gravitational potential.  For an atmosphere in complete equilibrium ($\mu = \mathrm{const}$, $T = \mathrm{const}$), calculate the pressure as a function of position, $P = P(z)$, and show that it agrees with considerations from hydrostatic balance, equation~(\ref{e.planar-hydrostatic}).

\item 
\begin{enumerate}
\item Solve equation~(\ref{e.saha-reduced}) for a density of $10^{16}\nsp\cm^{-3}$ (a fiducial value for the solar photosphere) and find the half-ionization temperature, i.e., the temperature at which $x=0.5$.  Explain the reason for the discrepancy between the half-ionization temperature and $\kB T = 13.6\nsp\eV$.

\item At this half-ionization temperature, what is the occupancy of the excited levels of the hydrogen atom?  Do we need to worry about corrections to the ionization from these excited states?
\end{enumerate}

\item Show that equation~(\ref{e.plasma-parameter}) is equivalent to $\Gamma \ll 1$ for a single species plasma.

\item Show that for a non-relativistic plasma, the magnetic interaction between two charged particles is much less than the electrostatic interaction.

\item Show that the net charge in the shielding cloud about an ion of charge $Ze$ is $-Ze$; the shielding cloud cancels out the ion's charge.

\item 
\begin{enumerate} 
\item\label{p.zero-pressure-iron} In the zero-temperature limit (electrons are fully degenerate) use the charge-neutral sphere approximation (p.~\pageref{e.madelung-total}) to calculate the density at which completely ionized \iron[56] has zero pressure.

\item Estimate the pressure that would be required to compress \iron[56] at the density found in part~\ref{p.zero-pressure-iron}.
\end{enumerate}

\item Now write the total pressure as the sum of electron and Coulomb pressure, and use the virial scalings that we derived for pressure and density in terms of mass and radius, to obtain a relation between mass and radius for a cold object (``a rock'').  Find the mass having the largest radius, and express this mass in terms of fundamental physical constants.  How does it compare with the mass of Jupiter?  Scale the mass and radii to that of Jupiter, and plot $R(M)$ for pure \hydrogen, pure \helium, and pure \carbon\ objects.  Also indicate on this plot the masses and radii of the Jovian planets for comparison.

\end{enumerate}
